---
title: "New_scrnas"
output: pdf_document
---
THe necessary packages
```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(Rsubread) #for feature counting, count matrix creation
library(SingleCellExperiment)
library(data.table)
library(org.Hs.eg.db)
library(annotate) #for the gene ids
library(biomaRt)
library(scRNAseq)
library(scater)
library(scran)
library(bluster)
library(AllelicImbalance)
library(XCIR)
library(DESeq2)
library(reshape2)
library(dplyr)
library(scds)
library(edgeR) 
library(rsvd)
library(Rtsne)
library(cowplot)
```
## Data loading

```{r cars}

test <- c("W11","W117","W87","W99","W43","W103","W232p2","W159p2","W196p2","W330p2")
control <- c("W193D","W37D","W60D","W175D","W116D","W44D2","W300Dp2","W336Dp2","W404Dp2","W327Dp2")

cell_trimm <- function(x){
  print
  return(substr(x,17, (nchar(x)-11)))
}

w_2_103 <- read.csv("C:/Users/38598/Desktop/Xci/reports/batch2_features/W103_counts_matrix_new", row.names = 1)
w_2_11 <- read.csv("C:/Users/38598/Desktop/Xci/reports/batch2_features/W11_counts_matrix_new", row.names = 1)
w_2_37D <- read.csv("C:/Users/38598/Desktop/Xci/reports/batch2_features/W37D_counts_matrix_new", row.names = 1)
w_2_43 <- read.csv("C:/Users/38598/Desktop/Xci/reports/batch2_features/W43_counts_matrix_new", row.names = 1)
w_2_44D2 <- read.csv("C:/Users/38598/Desktop/Xci/reports/batch2_features/W44D2_counts_matrix_new", row.names = 1)
w_2_60D <- read.csv("C:/Users/38598/Desktop/Xci/reports/batch2_features/W60D_counts_matrix_new", row.names = 1)
w_2_87 <- read.csv("C:/Users/38598/Desktop/Xci/reports/batch2_features/W87_counts_matrix_new", row.names = 1)
w_2_99 <- read.csv("C:/Users/38598/Desktop/Xci/reports/batch2_features/W99_counts_matrix_new", row.names = 1)
w_2_116D <- read.csv("C:/Users/38598/Desktop/Xci/reports/batch2_features/W116D_counts_matrix_new", row.names = 1)
w_2_117 <- read.csv("C:/Users/38598/Desktop/Xci/reports/batch2_features/W117_counts_matrix_new", row.names = 1)
w_2_175D <- read.csv("C:/Users/38598/Desktop/Xci/reports/batch2_features/W175D_counts_matrix_new", row.names = 1)
w_2_193D <- read.csv("C:/Users/38598/Desktop/Xci/reports/batch2_features/W193D_counts_matrix_new", row.names = 1)


w_300 <- read.csv("C:/Users/38598/Desktop/Xci/reports/batch2_features/W300_counts_matrix_new", row.names = 1)
w_336 <- read.csv("C:/Users/38598/Desktop/Xci/reports/batch2_features/W336_counts_matrix_new", row.names = 1)
w_330 <- read.csv("C:/Users/38598/Desktop/Xci/reports/batch2_features/W330_counts_matrix_new", row.names = 1)
w_404 <- read.csv("C:/Users/38598/Desktop/Xci/reports/batch2_features/W404_counts_matrix_new", row.names = 1)
w_196 <- read.csv("C:/Users/38598/Desktop/Xci/reports/batch2_features/W196_counts_matrix_new", row.names = 1)
w_327 <- read.csv("C:/Users/38598/Desktop/Xci/reports/batch2_features/W327_counts_matrix_new", row.names = 1)
w_232 <- read.csv("C:/Users/38598/Desktop/Xci/reports/batch2_features/W232_counts_matrix_new", row.names = 1)
w_159 <- read.csv("C:/Users/38598/Desktop/Xci/reports/batch2_features/W159_counts_matrix_new", row.names = 1)

colnames(w_2_103) <- sapply(colnames(w_2_103), cell_trimm)
colnames(w_2_11) <- sapply(colnames(w_2_11), cell_trimm)
colnames(w_2_37D) <- sapply(colnames(w_2_37D), cell_trimm)
colnames(w_2_43) <- sapply(colnames(w_2_43), cell_trimm)
colnames(w_2_44D2) <- sapply(colnames(w_2_44D2), cell_trimm)
colnames(w_2_60D) <- sapply(colnames(w_2_60D), cell_trimm)
colnames(w_2_87) <- sapply(colnames(w_2_87), cell_trimm)
colnames(w_2_99) <- sapply(colnames(w_2_99), cell_trimm)
colnames(w_2_116D) <- sapply(colnames(w_2_116D), cell_trimm)
colnames(w_2_117) <- sapply(colnames(w_2_117), cell_trimm)
colnames(w_2_175D) <- sapply(colnames(w_2_175D), cell_trimm)
colnames(w_2_193D) <- sapply(colnames(w_2_193D), cell_trimm)

colnames(w_300) <- sapply(colnames(w_300), cell_trimm)
colnames(w_327) <- sapply(colnames(w_327), cell_trimm)
colnames(w_336) <- sapply(colnames(w_336), cell_trimm)
colnames(w_232) <- sapply(colnames(w_232), cell_trimm)
colnames(w_404) <- sapply(colnames(w_404), cell_trimm)
colnames(w_196) <- sapply(colnames(w_196), cell_trimm)
colnames(w_159) <- sapply(colnames(w_159), cell_trimm)
colnames(w_330) <- sapply(colnames(w_330), cell_trimm)

experiment1 <- cbind(w_2_103,w_2_11,w_2_37D,w_2_43,w_2_44D2,w_2_60D,w_2_87,w_2_99,w_2_116D,w_2_117,w_2_175D, w_2_193D)
experiment2 <- cbind(w_330,w_300,w_327,w_336,w_404,w_232,w_196,w_159)


cdata1 <- data.frame(cell_names=colnames(experiment1))
cdata2 <- data.frame(cell_names=colnames(experiment2))
cdata1$patient <- sub("_[^_]+$", "",cdata1$cell_names)
cdata2$patient <- sub("_[^_]+$", "",cdata2$cell_names)

cdata1$type <- ""    
cdata1[cdata1$patient%in%control,]$type <- "control"
cdata1[cdata1$patient%in%test,]$type <- "test"
cdata2$type <- ""    
cdata2[cdata2$patient%in%control,]$type <- "control"
cdata2[cdata2$patient%in%test,]$type <- "test"

anot <- as.data.table(toTable(org.Hs.egCHR))
setnames(anot, "gene_id", "id")
chr_anot <- function(x){
  return(anot[id==x, chromosome])
}
chr_name1 <- sapply(rownames(experiment1), chr_anot) 
chr_name2 <- sapply(rownames(experiment2), chr_anot) 

chr1 <- as.data.frame(data.table(chr_name1))
chr2 <- as.data.frame(data.table(chr_name2))


gene_name1 <- sapply(rownames(experiment1), lookUp,'org.Hs.eg', "SYMBOL")
gene_name2 <- sapply(rownames(experiment2), lookUp,'org.Hs.eg', "SYMBOL")

ensembl_name1 <- sapply(rownames(experiment1), lookUp,'org.Hs.eg', "ENSEMBL")
ensembl_name2 <- sapply(rownames(experiment2), lookUp,'org.Hs.eg', "ENSEMBL")
```

Warning: There was a change in how Mitochondrial genes are named in all of these databases, they now omit the "Mt-" part of the name, making some of these workflows and functions incorrectly result in no Mt genes detected: fixing below by using a list of mt gene names, including non protein coding ones just in case

```{r}
mito_genes <- c("ATP6","ATP8","CYB","ND4L",
                "ND1","ND2","ND3","ND4",
                "ND5","ND6","RNR1","RNR2",
                "TD","TE","TF","TG","TH",
                "TI","TK","TL1","TL2","TM",
                "TN","TP","TQ","TR","TS1",
                "TS2","TV","TW","TY")
```


using the single cell experiment class

```{r}
sce1 <- SingleCellExperiment(assays=list(counts = experiment1), colData = cdata)
sce2 <- SingleCellExperiment(assays=list(counts = experiment2), colData = cdata)
rowData(sce1)$chr <- chr_name1 
rowData(sce2)$chr <- chr_name2 


rowData(sce1)$geneid <- rownames(sce1)
rowData(sce2)$geneid <- rownames(sce2)

rowData(sce1)$Symbol <- gene_name1
rowData(sce2)$Symbol <- gene_name2

rowData(sce1)$Ensembl <- ensembl_name1
rowData(sce2)$Ensembl <- ensembl_name2

rownames(sce1) <- gene_name1
rownames(sce2) <- gene_name2

head(rowData(sce1))
head(colData(sce1))
head(rowData(sce2))
head(colData(sce2))

```


Implementing the extra quality controll
 Quality control filtered out cells with >5% mitochondrial genes and either very low (<200) or very high (3 times the median absolute deviation of genes) number of genes.

```{r}
sceqc <- sce1
is.mito <- rowData(sceqc)$Symbol%in%mito_genes
stats <- perCellQCMetrics(sceqc, subsets=list(Mito=is.mito), threshold=5) 
high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
sum(high.mito)
sum(stats$subsets_Mito_percent>5)



colData(sceqc) <- cbind(colData(sceqc), stats)
sceqc$high_mito <- high.mito
sceqc$five_percent_mito <- stats$subsets_Mito_percent>5


plotColData(sceqc, x="sum", y="subsets_Mito_percent",
        colour_by="high_mito") + scale_x_log10()

plotColData(sceqc, x="detected", y="subsets_Mito_percent",
        colour_by="five_percent_mito") + scale_x_log10() +xlab("Number of detected features")+ylab("% counts belongiing to mitochondrial genes")  + guides(fill=guide_legend(title="New Legend Title"))


plotColData(sceqc, x="sum", y="subsets_Mito_percent",
        colour_by="patient") + scale_x_log10()
plotColData(sceqc, x="sum", y="subsets_Mito_percent",
        colour_by="plate") + scale_x_log10()

table(sceqc[,sceqc$five_percent_mito]$patient)


discarded <- sceqc[,sceqc$five_percent_mito] 
sceqc <- sceqc[,!sceqc$five_percent_mito] 

```

Filtering by counts being too high or too low as well

```{r}

sum(stats$detected<200)  



m <- mad(stats$detected)
sum(stats$detected>(3*m))


stats$mad <- abs(stats$detected-median(stats$detected))
stats$mad_over_3 <- stats$detected>(3*m)
stats$mad_over_3_b <- stats$detected>(median(stats$detected)+(3*m))
colData(sceqc) <- cbind(colData(sceqc), stats)
plotColData(sceqc, x="detected", y="mad",
        colour_by="mad_over_3") + scale_x_log10()

plotColData(sceqc, x="detected", y="mad",
        colour_by="mad") + scale_x_log10()


plotColData(sceqc, x="detected", y="subsets_Mito_percent",
        colour_by="mad_over_3") + scale_x_log10()+xlab("Number of detected features")+ylab("% counts belonging to mitochondrial genes")



plotColData(sceqc, x="detected", y="sum",
        colour_by="mad_over_3_b") + scale_x_log10()+xlab("Number of detected features")+ylab("read count")



```

Normalization and clustering 

```{r}

sceqc <- scuttle::logNormCounts(sceqc)

set.seed(94)

sceqc <- runUMAP(sceqc, scale=T, ntop =500)

plotUMAP(sceqc,colour_by=I(sceqc$type), ncomponents = 2)
plotUMAP(sceqc,colour_by=I(sceqc$patient), ncomponents = 2)
plotUMAP(sceqc,colour_by=I(sceqc$plate), ncomponents = 2,shape_by=I(sceqc$type))


```
Analyzing UMAPS, top variance genes, get fold changes

```{r}
a1 <- getTopHVGs(sceqc)
a2 <- getTopHVGs(sceqc, row.names = F) 

hvg_table <- rowData(sceqc[a2[1:500],])

colLabels(sceqc) <- clusterCells(sceqc, use.dimred='UMAP',  BLUSPARAM=KmeansParam(centers=4) )
plotUMAP(sceqc,colour_by="label", ncomponents = 2, text_by="label")

markers <- findMarkers(sceqc,test.type = "wilcox",pval.type = "all")

markers2 <- findMarkers(sceqc,test.type = "t", pval.type = "all")
markers3 <- findMarkers(sceqc,test.type = "t", pval.type = "any")



markers_test <- findMarkers(sceqc, groups = sceqc$type, test.type = "wilcox", pval.type="all")
markers_test$test


markers_test <- findMarkers(sceqc[,colLabels(sceqc)==4], groups = sceqc[,colLabels(sceqc)==4]$type, test.type = "wilcox",pval.type = "all")
markers_test$test

write.csv(hvg_table,"C:/Users/38598/Desktop/hvg_top500.csv")
```
